Scheduling Musical Events in Max/MSP with Scheme for Max

### 2021-06-25 paper-1/paper-1.tex
saving

As each visual box in a patcher has state that is retained between messages, we can see that when
programming the visual patcher (a.k.a. ``patching"), we are in effect programming with
object instantiations rather than classes. In fact, in the Max engine, the act of creating
a new visual-box and placing it on the patcher canvas does indeed instantiate an object, creating
a fresh copy of the prototype's data structure and adding it to a graph of other objects.
This is in contrast to computer music languages such as CSound, where the user programs instruments 
with functions and the engine creates an instantiated data structure on each note-event sent to 
the instrument from a score - in effect the instruments act as object builders and note-events 
become objects that exist for the duration of the note played.

% TODO: cite Dannenburg on this point  %

%% done %%
%% perhaps cite SDK documentation? or eric lyon book? %%

@@

However, Scheme for Max is unique in bringing a fully-fledged Scheme interpreter
to the Max environment as a first-class Max object. This is significant in that Max
is arguably the most widely used truly programmable
music platform, and certainly the most widely deployed through its use in 
the very successful Ableton Live digital audio workstation, as Max For Live.

@@

%% figure of pipe use
%% figure of naive solution

@@

The \textbf{js} object in Max is not an ideal solution to our problem for several reasons. 
The most serious is that it is limited to running only in the Max low-priority GUI thread \cite{Cycling74}.
As this thread is also used for Max file i/o and graphic redrawing, the result is that latency
is potentially high and indeterminate, meaning musical timing accuracy is unpredictable and often poor.
The human ear is very sensitive to time, and perceives events separated by 20 ms as discrete
events. 
% TODO: maybe a citation here??
Delays of tens of milliseconds are enough to sound like errors in playback
of highly rhythmic music. This means that it takes little other activity in the
GUI thread to delay our scheduled events enough to be audibly incorrect. 
For some purposes, this is acceptable, but for the creation of accurate 
musical sequencers or algorithmic music engines, the \textbf{js} object is unreliable.

@@

% TODO figure of future events

@@

The scheduling system describe does have several limitations worth noting.
The s7 Scheme implementation includes a tracing garbage collector 
that must run occasionally, pausing other processing \cite{Matheussen2020}.
The garbage collector can, however, be paused and run on demand. Tests on the author's
system, with the garbage collector forced to run every 100 ms to prevent
significant build up, showed the garbage collection taking an average of 1 to 2 ms per pass
when run in a substantial application (a 16 track sequencer written entirely in S4M
with the computer also generating audio for the tracks with commercial
software synthesizers).  
The audio rendered was recorded and examined
examined in a digital audio workstation to check for timing discrepancies. The
results of these preliminary tests were that timing was unaffected by the 
garbage collector so long as Max was configured with 
a large enough output buffer to allow the collector to run within the 
latency period of this buffer.
Using Max with an output buffer of 256 samples produces an output latency
of approximately 6-8 ms (depending on the Max signal vector as well), which
was large enough for the collector to run without issue, but is 
low enough to be acceptable for real time performance. With software synthesis
loads at closer to the maximum possible on the author's computer,
an output buffer of 512 samples (~11 ms output latency) was necessary to prevent
any timing errors. This is encouraging as these buffer sizes are not atypical
for commercial music production on digital audio workstations and are widely
considered as being in acceptable ranges for real-time music systems \cite{Brandt1998}.

In addition to output latency, there is the factor of internal latency.
Max, like many computer music programming environments, generates audio 
in vectors of samples, with
each vector generated from one processing pass, and (normally) control
rate passes conducted also a maximum of once per audio vector pass. 
% TODO could add citation here
The size of this signal vector thus has a large effect on performance, with larger
vectors requiring significantly less CPU use. As audio rendering of note onsets originating
from control messages (the normal scenario) must thus be aligned with 
signal vector boundaries, this produces delays of note onsets by up to
the time period of one vector, a phenomenon known as ``timing jitter" in the nomenclature.
However, the more recent scheduling facilities in Max, including the 
clock facilities used by S4M, are designed to compensate for this jitter,
maintaining long term temporal accuracy despite note onset delays. The mechanism
by which this is done is unfortunately not publically documented or available
in open-source code, but one can assume that it is some variant of the common pattern 
whereby scheduled events are aware of when they are supposed to be running
(``logical-time"), and if they in turn schedule subsequent events, the time for these subsequent
events will be shifted to compensate for the jitter, based 
on the logical time of the scheduling event, rather than its
actual (post-jitter) real time \cite{Anderson1986}.
The author also conducted personal tests of 
this compensation, comparing both high and low signal vector sizes to produce
variable delays in note onsets, and it was determined
that so long as the overall output latency is again large enough to allow
the GC to run in time, self-scheduling events in a long piece do stay
accurate in that jitter does not accumulate over the piece - 
note onsets stay accurate to within the delay of one signal vector of samples,
regardless of the length of the piece of music.

>>>>

As each visual box in a patcher has state that is retained between messages, we can see that when
programming the visual patcher (a.k.a. ``patching"), we are in effect programming with
object instantiations rather than classes. In fact, in the Max engine, the act of creating
a new visual-box and placing it on the patcher canvas does indeed instantiate an object, creating
a fresh copy of the prototype's data structure and adding it to a graph of other objects.
This is in contrast to computer music languages such as CSound, where the user programs instruments 
with functions and the engine creates an instantiated data structure on each note-event sent to 
the instrument from a score - in effect the instruments act as object builders and note-events 
become objects that exist for the duration of the note played \cite{Lazzarini2013}.

@@

However, Scheme for Max is unique in bringing a fully-fledged Scheme interpreter
to the Max environment as a first-class Max object, implemented and running
identically to any built-in object. This is significant in that Max
is arguably the most widely used truly programmable
music platform, and certainly the most widely deployed through its use in 
the very successful Ableton Live digital audio workstation, as Max For Live.

@@

\begin{figure}[H]
  \centering
  \includegraphics[width=.5\linewidth]{fig-5-pipe}
  \caption{Delaying a list message with the pipe object}
  \Description{.}
\end{figure}

@@

The \textbf{js} object in Max is not an ideal solution to our problem for several reasons. 
The most serious is that it is limited to running only in the Max low-priority GUI thread \cite{Cycling74}.
As this thread is also used for Max file i/o and graphic redrawing, the result is that latency
is potentially high and indeterminate, meaning musical timing accuracy is unpredictable and often poor.
The human ear is very sensitive to time, with delays of tens of milliseconds enough to sound like errors in playback
of highly rhythmic music. This means that it takes little other activity in the
GUI thread to delay our scheduled events enough to be audibly incorrect. 
For some purposes, this is acceptable, but for the creation of accurate 
musical sequencers or algorithmic music engines, the \textbf{js} object is unreliable.

@@

\begin{figure}[H]
  \centering
  \includegraphics[width=.5\linewidth]{fig-6-s4m-delay}
  \caption{Delaying a mixed list message, with dial input as is and as will be}
  \Description{.}
\end{figure}

@@

The scheduling system describe does have several limitations worth noting.
The s7 Scheme implementation includes a tracing garbage collector 
that must run occasionally, pausing other processing \cite{Matheussen2020}.
The garbage collector can, however, be paused and run on demand. Tests on the author's
system, with the garbage collector forced to run every 100 ms to prevent
significant build up, showed the garbage collection taking an average of 1 to 2 ms per pass
when run in a substantial application (a 16 track sequencer written entirely in S4M
with the computer also generating audio for the tracks with commercial
software synthesizers).  
The audio rendered was recorded and examined
examined in a digital audio workstation to check for timing discrepancies. The
results of these preliminary tests were that timing was unaffected by the 
garbage collector so long as Max was configured with 
a large enough output buffer to allow the collector to run within the 
latency period of this buffer.
Using Max with an output buffer of 256 samples produces an output latency
of approximately 6-8 ms (depending on the Max signal vector as well), which
was large enough for the collector to run without issue, but is 
low enough to be acceptable for real time performance. With software synthesis
loads at closer to the maximum possible on the author's computer,
an output buffer of 512 samples (approx. 11 ms output latency) was necessary to prevent
any timing errors. This is encouraging as these buffer sizes are not atypical
for commercial music production on digital audio workstations and are widely
considered as being in acceptable ranges for real-time music systems \cite{Brandt1998}.

In addition to output latency, there is the factor of internal latency.
Max, like many computer music programming environments, generates audio 
in vectors of samples, with
each vector generated from one processing pass, and (normally) control
rate passes conducted also a maximum of once per audio vector pass \cite{Puckette2002}.
The size of this signal vector thus has a large effect on performance, with larger
vectors requiring significantly less CPU use. As audio rendering of note onsets originating
from control messages (the normal scenario) must thus be aligned with 
signal vector boundaries, this produces delays of note onsets by up to
the time period of one vector, a phenomenon known as ``timing jitter" in the nomenclature.
However, the more recent scheduling facilities in Max, including the 
clock facilities used by S4M, are implemented to compensate for this jitter,
maintaining long term temporal accuracy despite note onset delays. The mechanism
by which this is done is unfortunately not publically documented or available
in open-source code, but one can assume that it is some variant of the common pattern 
whereby scheduled events are aware of when they are supposed to be running
(``logical-time"), and if they in turn schedule subsequent events, the time for these subsequent
events will be shifted to compensate for the jitter, based 
on the logical time of the scheduling event, rather than its
actual (post-jitter) real time \cite{Anderson1986}.
The author also conducted personal tests of 
this compensation, comparing both high and low signal vector sizes to produce
variable delays in note onsets, and it was determined
that so long as the overall output latency is again large enough to allow
the GC to run in time, self-scheduling events in a long piece do stay
accurate in that jitter does not accumulate over the piece - 
note onsets stay accurate to within the delay of one signal vector of samples,
regardless of the length of the piece of music.
